Just Take the Bus: Analysis of Toronto Bike Theft from 2022-2024

### 2024-11-11 Paper/Article.Rmd
added intro and top 10 neighbourhood map

library(ggplot2)
library(maps)
library(dplyr)
library(tidyverse)
library(sf)
library(mapdata)
library(ggspatial)
library(opendatatoronto)
library(here)
```

@@

bike_thefts <- read_csv(here("Data", "Bicycle_Thefts_Open_Data.csv")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
bike_thefts_sf <- read_sf(here("Data", "Bicycle_Thefts_Open_Data.geojson")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
```

@@

## TIME TRENDS ANALYSIS

@@

### Neighbour hood analysis and Maps

@@

### Bike Theft by Type and Cost
Here we com[are] the frequency of thefts by bike type and cost distribution:

@@

```{r}
# Box plot for bike cost distribution by status
ggplot(bike_thefts %>% filter(!is.na(BIKE_COST)), aes(x = STATUS, y = BIKE_COST)) +
  geom_boxplot() +
  labs(title = "Bike Cost Distribution (Recovered vs Stolen)", x = "Status", y = "Cost ($)") +
  theme_minimal()
```
### Reporting Delays
Analyzing the time delay between theft occurrence and reporting date (Not working, got to tweajk it some more)

```{r}
#| warning: false
#| message: false
# calcuate reporting delay
bike_thefts <- bike_thefts %>%
  mutate(reporting_delay = as.numeric(difftime(ymd_hms(REPORT_DATE), ymd_hms(OCC_DATE), units = "days")))

@@

### Theft Recovery Analysis by Neighborhood
which neighborhoods have higher recovery rates and analyzing the recovered vs unrecovered bikes by neighborhood

@@

>>>>

library(ggplot2)
library(dplyr)
library(tidyverse)
library(sf)
library(mapdata)
library(ggspatial)
library(opendatatoronto)
library(here)
```

@@

bike_thefts <- read_csv(here("Data", "Bike_thefts_raw.csv")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
bike_thefts_sf <- read_sf(here("Data", "Bicycle_Thefts_Open_Data.geojson")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
```

@@

Many cities are promoting bicycles to reduce reliance on cars,
recognizing significant benefits cycling brings to the environment,
public health and economy. As cyclists continue to gain popularity in
Toronto, bicycle theft has emerged as a growing concern leaving many
feeling vulnerable. With the increasing number of cyclists on the road,
knowing when and where bike thefts are most likely to occur is crucial
for protecting valuable bicycles. This analysis aims to inform the
general public and cyclists about neighborhoods with the highest rate of
thefts and times when their bikes are most likely to be stolen. This
would not only raise awareness but also help people make more informed
decisions to secure their bikes, ultimately reducing the risk of theft
and fostering a safer cycling environment in Toronto.

**Top 10  neighborhoods that  contribute to the most number of bicycle
thefts across Toronto ?**

We can see the worst neighborhood for bike thefts in Toronto is
Yonge-Bay Corridor followed by Downtown Yonge East and St Lawrence East
Bayfront, with each exceeding 1000 incidents from 2022 to 2024. In the
past three years, about 5100 bikes were reported stolen from these
areas. These neighborhoods contribute to approximately 17% of total
bicycle thefts reported across Toronto. This high concentration reflects
a combination of factors such as lack of secure bike storage, crowded
public spaces making these areas prime targets for thieves.

How much do they contribute to total theft from 2022 to 2024 ? \
(Comparing with premise type)An in-depth analysis of premise type
unfolds that bicycles last parked outside building premises, whether
residential or commercial are particularly vulnerable to theft. This
trend is consistently observed in top 10 neighborhoods for bike theft.
These locations, especially because of high foot or bike traffic,
provide easy access and quick escape routes for criminals.

\

**When is your bike most likely to be stolen(day of the week, month) ?**

\

Based on the figures below, it is pretty evident the highest frequency
of bicycle thefts occurs during July and August. This aligns with the
warmer summer months when more people tend to ride bicycles, increasing
both bicycles in public spaces and likelihood of theft. Thefts begin to
build up during spring, peaking in summer, followed by a steady decline
through fall and winter. The data suggests thefts are more common during
weekdays, potentially because most people use bicycles to commute to
work or school. Bicycles parked during weekdays are often unattended for
several hours, providing thieves with ample time to steal them.

Additionally, the majority of bicycles get stolen during the night than
in morning, afternoon or evening.

@@

### Neighbourhood analysis and Maps

@@

### Bike Theft by Type and Cost

Here we com[are] the frequency of thefts by bike type and cost
distribution:

@@

```{r}
# Box plot for bike cost distribution by status
ggplot(bike_thefts %>% filter(!is.na(BIKE_COST)), aes(x = STATUS, y = BIKE_COST)) +
  geom_boxplot() +
  labs(title = "Bike Cost Distribution (Recovered vs Stolen)", x = "Status", y = "Cost ($)") +
  theme_minimal()
```

### Reporting Delays

Analyzing the time delay between theft occurrence and reporting date
(Not working, got to tweajk it some more)

```{r}
#| include: FALSE
#| warning: false
#| message: false
# calcuate reporting delay
bike_thefts <- bike_thefts %>%
  mutate(reporting_delay = as.numeric(difftime(ymd_hms(REPORT_DATE), ymd_hms(OCC_DATE), units = "days")))

@@

### Theft Recovery Analysis by Neighborhood

which neighborhoods have higher recovery rates and analyzing the
recovered vs unrecovered bikes by neighborhood

@@

```{r}
# map

# Count and filter top 10 neighborhoods with the highest thefts
top_neighborhoods <- bike_thefts %>%
  count(NEIGHBOURHOOD_158) %>%
  slice_max(n, n = 10)

# Treemap plot
treemap::treemap(
  top_neighborhoods,
  index = "NEIGHBOURHOOD_158",  # Using neighborhood names as the index
  vSize = "n",                  # Size by count of thefts
  vColor = "n",                 # Color by count of thefts
  palette = c("#2b5c8a", "goldenrod1", "#9e3d22"),
  type = "manual",
  align.labels = list(c("right", "top"), c("left", "bottom")),
  overlap.labels = 0.2,
  title = "Number of Bicycle Thefts by Neighborhood",
  fontsize.title = 18,
  title.legend = "Frequency",
  fontsize.labels = 10, 
)
```
